Supplementary Information

### 2021-07-26 doc/manuscript/response_to_reviews.Rmd

---
title: ''
output: 
    word_document
bibliography: references.bib
csl: global-change-biology.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Response to Reviewer(s)' Comments to Author:

## Reviewer: 1

The research is overall a worthwhile topic regarding joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests.
I found the paper to be well-written and I appreciate the work you have accomplished. I consider the approach combining simultaneously the different factors controlling tree growth (non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year) highly appropriate for exploring species and stand responses to global changes in different biomes.
The study is well performed and the results and their interpretations are sound. The results are convincing and tree-ring data are nicely discussed in the context of global changes. However, given the complexity of the results, it would be very useful to report a summary picture of the main results by biome/ecoregions and level of naturalness of the studied stand (old-growth, seminatural forest ecosystems or secondary forests, reforestation/afforestation if present).

I have a few minor remarks that the authors can consider:

1. Lines 178-180 please insert citations for this statement. I suggest also consider artefact of sampling biases caused by the absence of old, fast-growing trees (cf. “slow-grower survivorship bias”) and of young, slow-growing trees (cf. “big-tree selection bias”) in the dataset ( see Duchesne, L., Houle, D., Ouimet, R., Caldwell, L., Gloor, M., & Brienen, R. (2019). Large apparent growth increases in boreal forests inferred from tree-rings are an artefact of sampling biases. Scientific reports, 9(1), 1-9.).

Lines 180-182 please quote at least one supporting reference.

Lines 646 -648 in temperate closed old-growth forests an opposite behaviour is often reported for the oldest trees. Slow but increasing long-term growth allows for maximum longevity not only in beech, but also in oaks and pines   (see Piovesan, G., Biondi, F., Baliva, M., De Vivo, G., Marchiano, V., Schettino, A., & Di Filippo, A. (2019). Lessons from the wild: slow but increasing long‐term growth allows for maximum longevity in European beech. Ecology, 100(9), e02737 and references therein).

## Reviewer: 2

This article aims to better understand the relationship between tree-growth from tree-rings and both climate and tree size. While these are standard questions in forest ecology and dendrochronology, they are typically considered either independently or at the stand scale. These questions are inherently related (both climate and tree size influence growth), and this article considers them jointly. These do this by presenting a new approach that: 1) identifies climate variables that are most strongly related to three different growth metrics based on detrended ring-widths; 2) for each site and species combination tests a suite of models that have the raw growth metrics (not detrended) as the response, and some combination of the selected climate variables (from step 1), DBH, and the interaction of DBH and climate; and 3) estimates parameters for the model from step 2 with the lowest AIC. This process results in estimates of the relationship between each of the three growth metrics, and climate, size and DBH. Having a better understanding of these relationships, and their variability among sites and species, will ultimately help with prediction of future forest growth (by for example informing ecosystem models).

I commend the authors for preparing such a well-written document. Of all the articles I’ve reviewed, this one has the fewest typographical and grammatical errors (if any).
There are several questions/issues that should be clarified prior to publication. I think these are mostly minor, and likely will not require any further analysis (one comment may).

**Thank you!**

More general comments:

Data:
I am not sure why there is no data shown on any of the figures presented. Inclusion of the raw data should help support the results of the analysis. Since they are omitted, it makes me (as a reader) question if they are intentionally not shown as a result of some discrepancy. At the very least, I would like to see data included in a panel of each figure, and a comment as to why it was omitted in the rest.

Census:
I think the article mentions that most of these forest plots have repeat census data. How was this data used in the methods? If there were multiple census DBH measurements, how were these all used to constrain DBH time series?

Page 7-8:
RW, BAI, and deltaAGB are not independent, and BAI and delta ABH are not independent of DBH, which makes it somewhat complicated to interpret the difference in each of their relationships with climate and DBH. Further discussion on this interdependence is needed.
For example, there is reference to two relationships between RW and DBH. The first is that RW declines with age, and then second is that RW increase with DBH. Then the hypothesis that BAI peaks and then declines as a function of DBH. I don’t think it is possible (mathematically) to have RW increase with DBH be consistent with this BAI hypothesis.

 L222-224:
Dealing with tree and forest mortality is difficult. However, climate change is predicted to affect mortality in many forests. In turn, mortality can impact the growth of the remaining trees depending on the stature and location of the dead tree. This point should be addressed- the article presents the relationship between growth and climate only for the surviving trees. How does excluding growth data from the analysis potentially bias the results?

P19-20 (especially L301-303):
I think it could be made more clear that detrending was used only in the process of selecting strongly related climate variables, and not in the GLS models. Additionally, a comment on how the variability in the application of detrending method used here (thin plate splines) would impact climate variable selection would be warranted. There is a persistent question in dendrochronology about the amount of climate signal gets removed through the process of detrending.

L301-302: Motivation as to why one climate variable from each group was selected is missing. L302-303: DBH interactions with climate? Or climate-climate interactions?

P20-21: Various recent publications have highlighted the importance of the effect of climate interactions and climate disturbance interactions on tree growth. It is not clear to me why the DBH-climate interactions were included, but climate-climate interactions were not.

L313: From what I understand, a concave up fit is perfectly reasonable. For example, a standard exponential growth function is concave up, and is biologically possible. Clarify. Also clarify on how they were excluded.

L336-337: Is it also true that climate + DBH + climate x DBH + year is included? If not clarify why not.

L370: Clarify where the 20 comes from. Maybe mention it earlier after the discussion about what was included versus excluded. I tried to determine where this number came from but am lost.

Figure 3: There appear to be some concave up fits in here, which is not consistent with the statement that these were excluded.

Figure 4: This is really interesting. It looks like there is often little effect of climate on the growth of small trees, which seems expected but I’m not sure has been shown like this before.

Line 545-547: Agree that this is useful to help with prediction, but what is missing from the article is a more clear idea of how this could/would be done. What emerges here is that trees and forests are complicated. There seem to be some general patterns emerge, but for each of these there are always exceptions. How do you propose synthesizing all of this to actually improve predictions?

Section 590: It is really interesting to contract the variability between population scale and individual scale. There was a great paper by Jim Clark that constrasted these scales and showed that responses could appear different for each. Can’t remember the reference, but this seems to be the theme of this paper: individual scale can be quite different from population scale.

L630: I’m not sure “correct” is the right term. Since it is a joint model you can account for these things, but how this impacts the response is dependent on the other terms in the model. Maybe I missed something- is the climate signal removed before GLS modelling? I *think* the raw growth metrics are the response, right?

Very minor:

L94: enormous is a subjective word, change

L96-100: sentence awkward (not sure how the “the or to predict...” follows from first part of sentence)

L269: Defined ‘most important’

L280: linear and quadratic terms?

L308: So models with variables removed may include one climate variable and DBH? L311: Clarify “we ran every combination”. This is only for selected climate variable right? L318: Define “complete data set”

L344: “rarely differed significantly”; I think you mean your estimates when compared to the values obtained from the standard analysis (as opposed to your estimates differing from each other). Clarify.

Figure 2: It’s still not clear to me what “window open” and “window close” refer to. Clarify. L614: I’m still not sure what “cross-sectional” refers to. Does this mean census?

L627: What are you reconstructing? You mean back-calculating DBH? Probably a terminology thing.

L678, 683: Weird formatting issue in ref. Maybe journal software?

# References {-#References}

### 2021-07-26 doc/manuscript/ForestGEO_treerings.Rmd
quick pass through reviewer comments
went through all reviewer comments, commenting on how we'll address them. Made a few very small/quick fixes.

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002].
Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021] or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.

@@

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
Precipitation and temperature can have additive or interactive effects on growth [@foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

@@

Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021]. 
Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand.
Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation. 
Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.

@@

We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*). 
We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).

@@

For each species independently, we ran every combination of the candidate climate drivers and DBH, including both first- and second-order terms for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically.

@@

```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response, and relative strength of correlation, to climate variables across the range of potential seasonal windows. 
However, the seasonal window exhibiting the strongest climatic effect on growth, and even the most influential climate variable, sometimes differed among the growth metrics.
For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., PET at SCBI Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Fig. `r climwin_HF_W`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

We found that interactions between climate drivers and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of total cases analyzed;  Table `r hypotheses`, Figs. `r interactions_fig`, `r RW_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure such as ours that can account for this pattern and other DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.

@@

Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig. `r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
First, we correct for changes in climate that may drive directional growth trends.
For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig. `r year_fig`).
Second, we show that growth rate -- by any metric -- varies nonlinearly with DBH and with patterns dependent upon the species and environmental context (Fig. `r size`), reinforcing the concept that growth trend analyses should incorporate cross-sectional analyses to correct for growth ontogeny [e.g., through regional curve standardization, @peters_detecting_2015].
Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
The latter addresses a third important challenge, as those growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
By including a random effect of tree, our approach likely reduces the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

@@

The rarity of positive growth trends observed here indicates that any growth benefit from elevated CO~2~ was outweighed by some combination of demographic changes and chronic environmental shifts.
This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], including previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017], although some studies have detected growth increases [e.g., @hember_tree_2019; @voelker_historical_2006].
A growth benefit of increasing atmospheric CO~2~ concentration is expected, based on physiological mechanisms, under water-limited conditions and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.

>>>>

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002].
Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2021].
This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the large uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.

@@

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
Precipitation and temperature can have additive or interactive effects on growth [@foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2021; @rollinson_climate_2021].

@@

Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2021; @takahashi_air_2020; @belmecheri_precipitation_2021]. 
Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2021].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand (**REF**).
Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation (**REF**). 
Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.

@@

We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*). 
We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).

@@

For each species independently, we ran every combination of the candidate climate drivers and DBH, including both first- and second-order terms for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.

@@

```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response, and relative strength of correlation, to climate variables across the range of potential seasonal windows. 
However, the seasonal window exhibiting the strongest climatic effect on growth, and even the most influential climate variable, sometimes differed among the growth metrics.
For eight of 20 site-variable group combinations (i.e., water and temperature, each at 10 sites), both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., PET at SCBI Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Fig. `r climwin_HF_W`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

We found that interactions between climate drivers and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of total cases analyzed;  Table `r hypotheses`, Figs. `r interactions_fig`, `r RW_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2021].
An analytical structure such as ours that can account for this pattern and other DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.

@@

Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig. `r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
First, we account for changes in climate that may drive directional growth trends.
For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig. `r year_fig`).
Second, we show that growth rate -- by any metric -- varies nonlinearly with DBH and with patterns dependent upon the species and environmental context (Fig. `r size`), reinforcing the concept that growth trend analyses should incorporate cross-sectional analyses to correct for growth ontogeny [e.g., through regional curve standardization, @peters_detecting_2015].
Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
The latter addresses a third important challenge, as those growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
By including a random effect of tree, our approach likely reduces the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

@@

The rarity of positive growth trends observed here indicates that any growth benefit from elevated CO~2~ was outweighed by some combination of demographic changes and chronic environmental shifts.
This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2021], including previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017], although some studies have detected growth increases [e.g., @hember_tree_2019; @voelker_historical_2006].
A growth benefit of increasing atmospheric CO~2~ concentration is expected, based on physiological mechanisms, under water-limited conditions and has been observed in young forests in experimental settings [@walker_integrating_2021]. 
However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2021], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
